---
title: "mumps"
author: "Jessica Leviton"
date: "4/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import}
library(tidyverse)
#mumps <- read.csv("mumps.csv", header=TRUE)

read_csv("mumps.csv") %>%
  select(week, state, reports=cases) -> mumps

mumps$Year <- substr(mumps$week, 1, 4)
mumps$weeknum <- substr(mumps$week, 5, 6)
mumps$time <- as.numeric(mumps$Year) + as.numeric(mumps$weeknum)/52

head(mumps)

#mumps_MI <- mumps[mumps$state == 'MI',]
mumps_MI <- subset(mumps, (state == 'MI'))
mumps_data <- subset(mumps_MI, (time > 1969.75) & (time < 1970.7), select=c("time", "reports"))


mumps_data_t0 <- mumps_data$time[1]


plot(mumps_data$reports~mumps_data$time, type='l')
```

Chapter 14 Notes:

```{r pomp}
library(pomp)


sir_step <- Csnippet("
double dN_SE = rbinom(S, 1-exp(-Beta*I/N*dt));
double dN_EI = rbinom(I, 1-exp(-mu_EI*dt));
double dN_IR = rbinom(I, 1-exp(-mu_IR*dt));
S -= dN_SE;
E += dN_SE - dN_EI; 
I += dN_EI - dN_IR;
H += dN_IR;
")

sir_init <- Csnippet("
S = nearbyint(eta*N);
E = 0;
I = 1;
H = 0;
")

dmeas <- Csnippet("
lik = dbinom(reports,H,rho,give_log);
")

rmeas <- Csnippet("
reports = rbinom(H,rho);
")


mumps_fixed_params <- c(N=9e6, mu_EI=21.28, mu_IR=52.14)
params <- c(Beta=200, rho=0.8, eta=0.5, mumps_fixed_params)

mumps_data %>%
pomp(
  times="time",
  obsnames="reports",
  t0=mumps_data_t0,
  params=params,
  rprocess=euler(sir_step, delta.t=1/52),
  rinit=sir_init,
  rmeasure=rmeas,
  dmeasure=dmeas,
  accumvars="H",
  partrans=parameter_trans(
    log=c("Beta"),
    logit=c("rho", "eta")
  ),
  statenames=c("S", "E", "I", "H"),
  paramnames=c("Beta", "mu_EI", "mu_IR", "eta", "rho", "N")
) -> mumpSIR



```

Now, we'll do the actual fitting

```{r local}
run_level <- 1
mumps_Np <- switch(run_level, 100, 1e3, 5e3)
mumps_Nmif <- switch(run_level, 10, 100, 200)
mumps_Nreps_eval <- switch(run_level, 2, 10, 20)
mumps_Nreps_local <- switch(run_level, 10, 20, 40)
mumps_Nreps_global <- switch(run_level, 10, 20, 100)
mumps_Nsim <- switch(run_level, 50, 100, 500)

library(foreach)
library(doParallel)
registerDoParallel()
library(doRNG)
registerDoRNG(3899882)

#Testing work:
# mumpSIR %>%
#   pfilter(Np=mumps_Np, params=params) -> pf
# 
# logLik(pf)
# plot(pf)

mumpSIR %>%
  simulate(params=params, nsim=10, format="data.frame", include.data=TRUE) -> y


y %>%
  ggplot(aes(x=time, y=reports, group=.id, color=factor(.id)))+
  geom_line()+
 # ylim(0,100)+
  scale_color_brewer(type="qual", palette=3)+
  guides(color=FALSE)

mumps_rw.sd_rp <- 0.02
mumps_rw.sd_ivp <- 0.2
mumps_cooling.fraction.50 <- 0.5
mumps_rw.sd <- rw.sd(
  Beta=mumps_rw.sd_rp, eta=ivp(mumps_rw.sd_rp), rho=mumps_rw.sd_rp
)

stew(file="results/mif.rda", {
  t2 <- system.time({
    m2 <- foreach(i=1:mumps_Nreps_local,
                  .packages='pomp', .combine=c) %dopar%
      mif2(mumpSIR,
           Np=mumps_Np,
           Nmif=mumps_Nmif,
           cooling.fraction.50=mumps_cooling.fraction.50,
           rw.sd=mumps_rw.sd)
    lik_m2 <- foreach(i=1:mumps_Nreps_local,
                      .packages='pomp', .combine=rbind) %dopar%
      logmeanexp(
        replicate(mumps_Nreps_eval, logLik(
          pfilter(mumpSIR, params=coef(m2[[i]]), Np=mumps_Np))),
        se=TRUE)
  })
})

coef(m2[[1]])
logLik(pfilter(mumpSIR, params=params, Np=mumps_Np))

r2 <- data.frame(logLik=lik_m2[,1], logLik_se=lik_m2[,2],
                 t(sapply(m2,coef)))

r2
summary(r2$logLik, digits=5)


```


```{r global}
mumps_box <- rbind(
  Beta=c(5, 5e6), eta=c(0, 0.4), rho=c(0.2, 0.9)
)

stew(file="results/box_eval.rda", {
  time_start_box_eval <- Sys.time()
  m3 <- foreach(i=1:mumps_Nreps_global, .packages='pomp',
      .combine=c) %dopar% mif2(m2[[1]],
      params=c(apply(mumps_box, 1, function(x)runif(1,x[1],x[2])),
               mumps_fixed_params)
      )
  
  lik_m3 <- foreach(i=1:mumps_Nreps_global, .packages='pomp',
                    .combine=rbind) %dopar% logmeanexp(
                      replicate(mumps_Nreps_eval,
                                logLik(pfilter(mumpSIR,
                                               params=coef(m3[[i]]), 
                                               Np=mumps_Np))),
                      se=TRUE)
  time_end_box_eval <- Sys.time()
})

r3 <- data.frame(logLik=lik_m3[,1], logLik_se=lik_m3[,2],
                 t(sapply(m3,coef)))

r3
```

Alright, cool. It looks like the "week" data is formated "YYYY##" where ## is the "number" of the week, ranging from 01 to 52.


This link could have some useful general information, but you should also check CDC and other sites for mumps info: https://www.cdc.gov/mmwr/preview/mmwrhtml/mm5513a3.htm


https://www.senate.michigan.gov/sfa/Economics/Michigan&USPopulation.PDF
According to michigan.gov, the population of Michigan only increased by about 2% from the year 1970 to the year 1974. I would argue that that's a small enough change that we can approximate the demographics as constant.
