---
title: "mumps"
author: "Jessica Leviton"
date: "4/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import, echo=FALSE, results=FALSE}
library(tidyverse)

read_csv("mumps.csv") %>%
  select(week, state, state_name, cases) -> mumps

mumps$Year <- substr(mumps$week, 1, 4)
mumps$weeknum <- substr(mumps$week, 5, 6)
mumps$time <- as.numeric(mumps$Year) + as.numeric(mumps$weeknum)/52

head(mumps)

#mumps_MI <- mumps[mumps$state == 'MI',]
#mumps_MI <- subset(mumps, (state == 'MI'))
#mumps_data <- subset(mumps_MI, (week > 197037) & (week < 197236), #select=c("time", "reports"))

mumps_data = mumps %>%
  filter(state_name == "MICHIGAN") %>%
  filter(week >= 197137 & week <= 197334) %>%
  select(cases) %>%
  mutate(week = 1:100) %>%
  relocate(week, cases)

head(mumps_data)
# mumps_data_t0 <- mumps_data$time[1]
mumps_data_t0 <- 0


plot(mumps_data$cases~mumps_data$week, type='l')
```

Mumps is a virus that was once a common cause of childhood hearing loss and one of the leading causes of military hospitalizations in the United States [SOURCE: CDC]. Following the licensing of the measles, mumps, and rubella (MMR) vaccine in 1971 and a successful national 2 dose vaccination program, the U.S. saw a 99% reduction in cases. [STILL CDC]

In this report, we examine whether mumps cases in the 1970s can be well modeled by an SEIR pomp model. We used data from Project Tycho, which collects infections disease data to be used for research purposes. We used weekly mumps reporting data from Michigan from September 1971 to September 1973.

Our model consists of 4 compartments, representing the Susceptible (S), those in a latency period (E), the Infected (I), and those that have Recovered (R) from the disease. Taking the simplest case by ignoring demography, we can compute the number of people in each compartment with the following fomulas:

$$
S(t) = S(0) - N_{SE}(t)
\\
E(t) = E(0) + N_{SE}(t) - N_{EI}(t)
\\
I(t) = I(0)
\\
R(t) = R(0) + N_{IR}(t)
$$

Chapter 14 Notes:

```{r pomp, cache=FALSE}
library(pomp)

# mumps_covar <- covariate_table(
#   t=mumps_data$week,
#   periodic.bspline.basis(t, nbasis=1,
#                          degree=3, period=1, names="xi%d"),
#   times="t"
# )

seir_step <- Csnippet("
double Beta;
Beta = exp(b1 + b2 * cos(M_2PI/52*t - Phi));
double dN_SE = rbinom(S, 1-exp(-Beta*I/N*dt));
double dN_EI = rbinom(E, 1-exp(-mu_EI*dt));
double dN_IR = rbinom(I, 1-exp(-mu_IR*dt));
S -= dN_SE;
E += dN_SE - dN_EI;
I += dN_EI - dN_IR;
H += dN_IR;
")

seir_init <- Csnippet("
S = nearbyint(eta*N);
E = 20;
I = 10;
H = 0;
")

# dmeas <- Csnippet("
# lik = dbinom(cases, H, rho, give_log);
# ")

dmeas <- Csnippet("
lik = dnbinom(cases, H, rho, give_log);
")

# rmeas <- Csnippet("
# cases = rbinom(H, rho);
# ")

rmeas <- Csnippet("
cases = rnbinom(H, rho);
")

mumps_data %>%
  pomp(
    times="week", t0=0,
    rprocess=euler(seir_step, delta.t=1/7),
    rinit=seir_init,
    rmeasure=rmeas,
    dmeasure=dmeas,
    accumvars="H",
    partrans=parameter_trans(
      #log=c("Beta", "mu_EI"),
      logit=c("rho", "eta")
    ),
    statenames=c("S", "E", "I", "H"),
    paramnames=c("b1", "b2", "Phi", "mu_EI", "mu_IR", "eta", "rho", "N")
  ) -> mumpSIR

mumps_fixed_params <- c(N=8881826, mu_EI=0.412, mu_IR=0.714)
params <- c(b1=1, b2=1, Phi=0.1, rho=0.8, eta=0.0216, mumps_fixed_params)

# params <- c(Beta=20, mu_EI=3.5, mu_IR=2, rho=0.5, eta=0.1, N=38000)
# fixed_params <- c(N=38000, mu_IR=2)

```

Now, we'll do the actual fitting

```{r local, cache=FALSE, echo=FALSE}
run_level <- 2
mumps_Np <- switch(run_level, 100, 1e3, 5e3)
mumps_Nmif <- switch(run_level, 10, 100, 200)
mumps_Nreps_eval <- switch(run_level, 2, 10, 20)
mumps_Nreps_local <- switch(run_level, 10, 20, 40)
mumps_Nreps_global <- switch(run_level, 10, 20, 100)
mumps_Nsim <- switch(run_level, 50, 100, 500)

library(foreach)
library(doParallel)
registerDoParallel()
library(doRNG)
registerDoRNG(3899882)

#Testing work:
# mumpSIR %>%
#   pfilter(Np=mumps_Np, params=params) -> pf
# 
# logLik(pf)
# plot(pf)

mumpSIR %>%
  simulate(params=params, nsim=10, format="data.frame", include.data=TRUE) -> y


y %>%
  ggplot(aes(x=week, y=cases, group=.id, color=factor(.id)))+
  geom_line()+
 # ylim(0,100)+
  scale_color_brewer(type="qual", palette=3)+
  guides(color=FALSE)


foreach(i=1:mumps_Nreps_local,.combine=c) %dopar% { 
  library(pomp)
  library(tidyverse)
  mumpSIR %>%
    mif2(
      params=params,
      Np=mumps_Np, Nmif=mumps_Nmif,
      cooling.fraction.50=0.5,
      rw.sd=rw.sd(b1=0.02, b2=0.02, Phi=0.02, rho=0.02, eta=ivp(0.02))
    )
} -> mifs_local

lik_local <- foreach(i=1:mumps_Nreps_local, .combine=rbind) %dopar% {
  library(pomp)
  library(tidyverse)
  logmeanexp(
    replicate(mumps_Nreps_eval,
      logLik(pfilter(mumpSIR,
        params=coef(mifs_local[[i]]), Np=mumps_Np))),
    se=TRUE)
}

lik_local
r_local <- data.frame(logLik=lik_local[,1], logLik_se=lik_local[,2],
                      t(sapply(mifs_local, coef)))

r_local
summary(r_local$logLik, digits=5)

# mifs_local %>% logLik() %>% logmeanexp(se=TRUE) -> L_pf
# L_pf

mifs_local %>%
  traces() %>%
  melt() %>%
  ggplot(aes(x=iteration,y=value,group=L1,color=factor(L1)))+
  geom_line()+
  guides(color=FALSE)+
  facet_wrap(~variable,scales="free_y")

```


```{r global, cache=FALSE}


mumps_box <- rbind(
  b1=c(0,5), b2=c(0,5), Phi=c(0, 2*pi), eta=c(0,0.10), rho=c(0, 0.9)
)

mifs_global <- foreach(i=1:mumps_Nreps_global, .packages='pomp', 
  .combine=c) %dopar% mif2(mifs_local[[1]],
    params=c(apply(mumps_box, 1, function(x)runif(1,x[1],x[2])),
    mumps_fixed_params)
  )

lik_global <- foreach(i=1:mumps_Nreps_global, .packages='pomp',
  .combine=rbind) %dopar% logmeanexp(
    replicate(mumps_Nreps_eval,
      logLik(pfilter(mumpSIR,
        params=coef(mifs_global[[i]]), Np=mumps_Np))),
    se=TRUE)

r_global <- data.frame(logLik=lik_global[,1], logLik_se=lik_global[,2],
                       t(sapply(mifs_global, coef)))

r_global <- r_global[order(-r_global$logLik),]
r_global
summary(r_global$logLik, digits=5)
```
```{r fit_sim}
fit_params <- r_global[1,]
fit_params$b1

best_params <- c(b1=fit_params$b1, b2=fit_params$b2, Phi=fit_params$Phi,
                 rho=fit_params$rho, eta=fit_params$eta, mumps_fixed_params)

mumpSIR %>%
  simulate(params=best_params, nsim=1, format="data.frame", include.data=TRUE) -> y
y

y %>%
  ggplot(aes(x=week, y=cases, group=.id, color=factor(.id)))+
  geom_line()+
 # ylim(0,100)+
  scale_color_brewer(type="qual", palette=3)+
  guides(color=FALSE)
```

Alright, cool. It looks like the "week" data is formated "YYYY##" where ## is the "number" of the week, ranging from 01 to 52.


This link could have some useful general information, but you should also check CDC and other sites for mumps info: https://www.cdc.gov/mmwr/preview/mmwrhtml/mm5513a3.htm


https://www.senate.michigan.gov/sfa/Economics/Michigan&USPopulation.PDF
According to michigan.gov, the population of Michigan only increased by about 2% from the year 1970 to the year 1974. I would argue that that's a small enough change that we can approximate the demographics as constant.

vaccination rate info: https://www.statista.com/statistics/385577/mmr-vaccination-rate-among-us-children-aged-19-35-months/

better vaccination rate info: https://www.cdc.gov/vaccines/pubs/pinkbook/downloads/appendices/g/coverage.pdf

https://kingaa.github.io/pomp/vignettes/oaxaca.html
